smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases

### 2023-08-04 manuscript/manuscript.qmd
first draft of main body text before shortening

```{r setup}
#| label: setup
#| include: false
library(cowplot)
library(here)
library(tidyverse)
library(readxl)
library(gt)
library(gtsummary)
library(flextable)
library(officer)
library(captioner)
library(smdi)
suppressPackageStartupMessages(library(ggimage))

@@

@@

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box). These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and not often performed [@carroll2020; @wood2004; @harel2012].

@@

### Patterns (taken from Van Buuren [@vanbuuren2018])

@@

### Mechanisms (taken from Sondhi and Weberpals et al. [@sondhi2023a])

@@

-   **MCAR**: The missingness depends and can be explained by other observed covariates.

@@

To overcome these shortcomings, we [@SentinelCI3] have recently developed and evaluated a principled approach combining a range of missing data diagnostics in an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle the partially observed confounder data (e.g., missing data imputation).

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi]. The goal of `smdi` is to provide a toolbox of different diagnostics to analyze the potentially underlying missingness patterns and mechanisms of partially observed confounders in RWE studies.

@@

The `smdi` R package was written in R language (version 4.2.1) employing the devtools [@devtools] and roxygen [@roxygen2] packages. The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi)>, Posit Public Package Manager (<https://packagemanager.posit.co/client/#/repos/2/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command. To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis. Additional resources such as documentation, articles and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

@fig-workflow illustrates the recommended workflow and functions to systematically approach diagnostics on partially observed confounders.

In general, the workflow comprises two broader diagnostic categories: descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, the parameters a user can specify, the returned results and the interpretation of results. Examples are illustrated using a simulated EHR dataset that is part of the package (more details can be found under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>). Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

To visually inspect potential missing data patterns, we re-exported the `gg_miss_upset()` function of the naniar package [@naniar-2]. This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all partially observed covariates [@ruddle2022]. For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both labs simultaneously (example @fig-examples b).

@@

The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate. According to Rubin's framework [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates. Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the partially observed covariate. If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.

To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with a corresponding range of minimum and maximum AMSDs observed. The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate. For a graphical visualization of this, the function returns a graph illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2] .

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate separately. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.

The function returns an object of class *rf* which generically prints an overview of the AUC values of all partially observed covariates. The AUC value is based on the prediction made in the respective test dataset. The *rf* object further returns a graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model. In case of inflated AUC values (\>0.9), the function prompts a message to the user detailing the most important predictor. If another partially observed covariate is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this covariate and run the diagnostics for each partially observed covariate independently.

@@

Finally, the third group diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the partially observed covariate and the outcome under study. The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset. In previous simulations, we observed characteristic patterns for univariate and adjusted associations of the missing indicator and the outcome that matched simulated underlying missingness mechanisms [@weberpals2023]. As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the partially observed covariate. Under MAR, given that missingness can be explained by observed covariates, a potential spurious association in the univariate model disappeared after adjustment. If the missingness followed any MNAR mechanism, an association was observed in both univariate and adjusted models.

Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not. The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each partially observed covariate.

@@

The `smdi_diagnose()` function enables a user to compute all of the three group diagnostics introduced above with just one function call.

```{r}
#| eval: false
#| echo: true
smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)"
  ) %>% 
  smdi_style_gt()
```

The function will return an object of class *smdi* containing a table with the results of all diagnostics for each specified partially observed covariate and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.

@@

Missing data are ubiquitous in real-world databases and may lead to bias and spurious results if not handled appropriately.

By comparing the resulting diagnostic parameters to expected estimates, based on previous simulation results [@weberpals2023], the diagnostics can provide valuable insights into underlying missingness structures and help elucidate if analytic approaches such as missing imputation analyses are viable options @fig-results.

@@

```{r}
#| label: fig-examples
#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively."
#| echo: false
#| warning: false
#| fig-dpi: 400
#| fig-width: 15
#| fig-height: 10
#| out-width: 110%

@@

```{r}
#| label: fig-results
#| eval: true
#| fig-cap: "Example of how smdi diagnostics can be applied to compute and compare diagnostic parameters of partially observed covariates to expected parameters of common missingness mechanisms based on a former large-scale simulation study [@weberpals2023]."
#| echo: false
#| out-width: 110%
knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))
```

>>>>

```{r setup}
#| label: setup
#| include: false
library(cowplot)
library(here)
library(tidyverse)
library(readxl)
library(gt)
library(captioner)
library(smdi)
library(ggimage)

@@

[**Article type:**]{.underline} Application Note

@@

In order to make an informed decision about the most appropriate analytic approach, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder (POC) data (see definitions box) [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms. These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the POC [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of such diagnostics is time-consuming, tedious and consequently not often performed [@carroll2020; @wood2004; @harel2012].

@@

### Patterns (adapted from Van Buuren [@vanbuuren2018])

@@

### Mechanisms [@sondhi2023a]

@@

-   **MAR**: The missingness depends and can be explained by other observed covariates.

@@

In light of these limitations, we have recently developed and evaluated a principled approach with multiple missing data diagnostics in an US EHR-claims database linkage [@weberpals2023]. The results of this large-scale study revealed that the combination of these diagnostics characterized missing data mechanisms well and provided helpful guidance for the appropriate choice of analytic methods to handle POC data (e.g., missing data imputation).

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi]. The goal of `smdi` was to provide a toolbox of different diagnostics to analyze the potentially underlying missingness patterns and mechanisms of POCs in RWE studies.

@@

The `smdi` R package was written in R language (version 4.2.1). The package is available on CRAN ([https://cran.r-project.org/web/packages/smdi](https://cran.r-project.org/web/packages/smdi)){.uri}) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command. To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis. Additional resources such as documentation and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

@fig-workflow illustrates the recommended workflow and functions to systematically approach diagnostics on POCs.

The workflow is generally categorized into descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, a selection of parameters a user can specify, the returned results and the results interpretation. Examples are illustrated using a simulated EHR dataset that is part of the package (more details under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>). For all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the POCs. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the POC(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

To visually inspect potential missing data patterns, we re-exported the `gg_miss_upset()` function of the naniar package [@naniar-2]. This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all POCs [@ruddle2022]. For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both labs simultaneously (example @fig-examples b).

@@

The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the POC. According to Rubin's framework [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates. Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the POC. If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.

To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs along with a corresponding range of minimum and maximum AMSDs observed. The object also returns a detailed 'Table 1' for each POC displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the POC. For a graphical visualization of this, the function returns a figure illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2].

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each POC, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each POC separately. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all POCs with the null hypothesis that the data is MCAR.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the POC given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.

The function returns an object of class *rf* which generically prints an overview of the AUC value of each POC. The AUC is based on the prediction made in the respective test dataset which is sampled as part of the function and for which the train-test split ratio can be chosen by the user (`train_test_ratio` parameter). The *rf* object further returns a graph for each POC displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model. In case of inflated AUC values (\>0.9), the function prompts a message to the user detailing the most important predictor. If another POC is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this POC and run the diagnostics for each POC independently.

@@

The third group diagnostics with the `smdi_outcome()` function examines the association of the missingness indicator of the POC and the outcome under study. The function will compute both a univariate model and a model adjusted for all other covariates included in the dataset. In previous simulations, we observed characteristic patterns for univariate and adjusted associations of the missing indicator and the outcome that matched simulated missingness mechanisms [@weberpals2023]. As one would expect, under a MCAR mechanism there was no difference in the outcome between patients with and without a value for the POC. Under MAR, given that missingness can be explained by observed covariates, a potential spurious association in the univariate model disappeared after adjustment. If the missingness followed any MNAR mechanism, an association was observed in both univariate and adjusted models.

Currently, `smdi_outcome()` supports three outcome regression types: linear regression (*lm* [@stats]) for continuous outcomes, logistic regression (*glm* [@stats]) for binary outcomes and a Cox proportional hazards model (*coxph* [@survival]) for time-to-event outcomes. Besides the specification of the regression type, the user needs to specify the name of the column that contains the outcome using the *form_lhs* parameter (e.g., `Surv(eventtime, status)` in case of a Cox model) and if resulting beta coefficients should be exponentiated or not. The function returns a table with univariate and adjusted beta coefficients and 95% confidence intervals for each POC.

@@

Finally, the `smdi_diagnose()` function enables a user to compute all three group diagnostics with just one function call.

```{r}
#| eval: false
#| echo: true

# minimal example of a smdi_diagnose() function call
smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3 # number of cores to parallelize computations on
  ) %>% 
  smdi_style_gt()
```

The function returns an object of class *smdi* containing a table with the results of all diagnostics for each specified POC and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.

@@

Missing data are ubiquitous in real-world databases and may lead to bias if not handled appropriately. We developed `smdi` to streamline routine diagnostics of missing data. By cross-checking the resulting diagnostic parameters to expected estimates, based on the outlined theoretical assumptions and previous simulation results [@weberpals2023], the diagnostics can provide valuable insights into underlying missingness structures and help elucidate if analytic approaches such as missing imputation analyses are viable options (example shown in @fig-results).

The package also comes with limitations, such that the true underlying missingness generating mechanism can never be inferred with absolute certainty from the observed data. Hence, it's important that diagnostic results are complemented with substantive expert knowledge to factor in how covariates are measured in routine care processes and contextualize potential reasons for missingness.

## Conclusions

The smdi package is a powerful and convenient tool to implement and carry out principled missing data diagnostics in RWE studies and improve the robustness of studies involving POCs by helping elucidate if certain analytic approaches are viable options for a given dataset.

@@

```{r}
#| label: fig-examples
#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively."
#| echo: false
#| warning: false
#| fig-dpi: 400
#| fig-width: 15
#| fig-height: 10

@@

```{r}
#| label: fig-results
#| eval: true
#| fig-cap: "Example of how smdi diagnostics can be applied to compute and compare diagnostic parameters of partially observed covariates to expected parameters of common missingness mechanisms based on a former large-scale simulation study [@weberpals2023]."
#| echo: false
knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))
```
